---
title: "Midterm Project"
output: html_document
author: Disheng Jiang
---
## Loading package for data cleaning
```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(patchwork)
```


# Problem 1

###### Import data
```{r import data}
nsr_df <- 
  janitor::clean_names(
  read_excel(
    path = "p8105_mtp_data.xlsx",
    skip = 8,
    na = c("NA", ".", "")) 
  )
```
Skip the first 8 rows of instructions 
Standardize column names using janitor::clean_names().

## Label Mapping
```{r initial data cleaning}
nsr_df <- nsr_df |> 
  mutate(
    sex = as.character(sex),
    eop_visibility_classification = as.character(eop_visibility_classification),
    eop_shape = as.character(eop_shape),
    sex_name = case_match(
      sex,
      "1" ~ "male",
      "0" ~ "female"
      ),
    eop_visi_class = case_match(
      eop_visibility_classification,
      "0" ~ "0 mm",
      "1" ~ "0-5 mm",
      "2" ~ ">= 5mm"
      ),
    eop_shape = replace_na(as.numeric(eop_shape), 0),
    age_group_name = case_match(
      age_group,
      "2" ~ "18-30", 
      "3" ~ "31-40",
      "4" ~ "41-50",
      "5" ~ "51-60",
      "6" ~ "60+",
      "7" ~ "60+",
      "8" ~ "60+",
      ),
    eop_size_mm = replace_na(as.numeric(eop_size_mm), 0),
    eop_size_group = case_match(
      eop_size,
      "0" ~ "0-5 mm",
      "1" ~ "5-10 mm",
      "2" ~ "10-15 mm",
      "3" ~ "15-20 mm",
      "4" ~ "20-25 mm",
      "5" ~ "25+ mm",
    ),
    fhp_category_group = case_match(
      fhp_category,
      "0" ~ "0-10mm",
      "1" ~ "10-20 mm",
      "2" ~ "20-30 mm",
      "3" ~ "30-40 mm",
      "4" ~ "40-50 mm",
      "5" ~ "50-60 mm",
      "6" ~ "60-70 mm"
    )
  )
```
Sex: 0→female, 1→male → sex_name (factor);

Type conversion: Convert eop_size_mm, fhp_size_mm, sex (0/1), and age_group (2–5; 6/7/8 -> 60+) to integers.

Age group: 2→18–30, 3→31–40, 4→41–50, 5→51–60, 6/7/8→60+ → age_group_name (ordered factor).

Maintain classification: Keep the author's binning rules

###### creating tables: 1
```{r show age sex percentage}
tab_sex <- nsr_df |> count(sex_name, name = "number_of_people") |> mutate(
  pct = round(100 * number_of_people / sum(number_of_people), 1)
  ) |>  
  knitr::kable(col.names = c("sex", "number_of_people", "pct (%)"))

tab_age_sex <- nsr_df |> 
  count(age_group_name, sex_name, name = "number_of_people") |>
  group_by(age_group_name) |> 
  mutate(
    pct = round(100 * number_of_people / sum(number_of_people),1)
    ) |>
  ungroup() |> 
  pivot_wider(
    id_cols = age_group_name,
    names_from = sex_name,                     
    values_from = c(number_of_people, pct),     
    names_glue = "{.value}_{sex_name}",         
    values_fill = list(number_of_people = 0, pct = 0)
  ) |>
  arrange(age_group_name) |> 
  mutate(
    total_number_of_people = number_of_people_female + number_of_people_male
  ) |> 
  select(age_group_name, total_number_of_people, number_of_people_female, pct_female, number_of_people_male, pct_male) |>   
  knitr::kable()

tab_sex
tab_age_sex
```

###### creating tables: 2
```{r show mean and sd for overall/age group}
fhp_overall <- nsr_df |> 
  summarise(
    n = sum(!is.na(eop_size_mm)),
    mean = mean(eop_size_mm, na.rm = TRUE),
    sd = sd(eop_size_mm, na.rm = TRUE)
    ) |> 
  knitr::kable()

fhp_by_age <- nsr_df |> 
  group_by(age_group_name) |>
  summarise(
    total_number_people = sum(!is.na(eop_size_mm)),
    mean = mean(eop_size_mm, na.rm = TRUE),
    sd = sd(eop_size_mm, na.rm = TRUE), .groups = "drop"
    ) 
fhp_by_age |> 
  knitr::kable()
fhp_overall
```

###### creating tables: 3
```{r show mean and sd for sex group}
sex_mean_sd <- nsr_df |> 
  group_by(sex_name) |> 
  summarise(
    mean = mean(fhp_size_mm, na.rm = TRUE),
    sd = sd(fhp_size_mm, na.rm = TRUE))

sex_mean_sd |> 
  knitr::kable(col.names = c("sex", "fhp_mean", "fhp_sd"))
sex_mean_sd
```
We analyze `r nrow(nsr_df)` participants, approximately balanced by sex within each age group. 

Core variables are: age/age_group, sex, continuous EOP (`eop_size_mm`) with its 0–5 category (`eop_size_code`), and continuous FHP (`fhp_size_mm`).

Issues: a small number of rows have inconsistencies between continuous measurements and their bins ( `eop_size_mm` vs `eop_size_code`).

# Problem 2
```{r define enlarged group}
# EEOP definition：Group EOP size 2–5 = Enlarged
df <- nsr_df |>
  mutate(
    eop_size = as.numeric(eop_size),       
    eop_enlarged = case_when(
      eop_size %in% c(0, 1) ~ "regular", 
      eop_size %in% c(2, 3, 4, 5) ~ "enlarged",
      TRUE ~ NA_character_                 
    )
  )
```

###### Improving Fig 3
```{r create left plot}
p_left <- df |> 
  filter(!is.na(fhp_size_mm), !is.na(age_group_name), !is.na(sex)) |> 
  ggplot(aes(x = age_group_name, y = fhp_size_mm, fill = sex_name)) +
  # box plot dipiction
  geom_boxplot(position = position_dodge(width = 0.8),
               outlier.alpha = 0.5, width = 0.7) +
  # plot a point at the mean of age & sex
  stat_summary(aes(group = sex_name),
               fun = mean, geom = "point",
               position = position_dodge(width = 0.8),
               shape = 23, size = 2.8, fill = "white", color = "black") +
  # Mark 1 digit-rounded mean
  stat_summary(aes(group = sex_name, label = sprintf("%.1f", after_stat(y))),
               fun = mean, geom = "text",
               position = position_dodge(width = 0.8),
               vjust = -0.6, size = 3) +
  labs(x = "Age group", y = "FHP size (mm)",
       title = "FHP distribution by age & sex") +
  ggplot2::scale_fill_viridis_d(name = "Sex") +
  # Leave a room to the title
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.10))) +
  theme_minimal(base_size = 12)
p_left
```

###### Improve Fig 4
```{r define prevalence}
prev <- df |> 
  filter(!is.na(eop_enlarged)) |> 
  summarise(
    n = n(),
    cases = sum(eop_enlarged == "enlarged"),
    prev = round(cases / n, 4),
    .by = c(age_group_name, sex_name)
  ) |> 
  arrange(age_group_name, sex_name) |> 
    filter(!is.na(age_group_name)) 
prev
```

###### Create right plot
```{r using ggplot}
p_right <- 
  ggplot(prev, aes(x = age_group_name, y = prev, fill = sex_name)) +
  geom_col(position = position_dodge(width = 0.8) , width = 0.7, alpha = 0.9) +
  # Link the line with the point
  geom_line(aes(color = sex_name, group = sex_name),
            position = position_dodge(width = 0.8) , linewidth = 1) +
  geom_point(aes(color = sex_name),
             position = position_dodge(width = 0.8) , size = 2) +
  # Mark the percentage label above
  geom_text(aes(label = scales::percent(prev, accuracy = 0.1),
                group = sex_name),
            position = position_dodge(width = 0.8) , vjust = -0.5, size = 3) +
  # Label y-axis with %
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Age Group",
       y = "Proportion with Enlarged EOP",
       title = "Enlarged EOP (EOP size 2–5) prevalence by age & sex") +
  ggplot2::scale_fill_viridis_d(name = "Sex") +
  ggplot2::scale_color_viridis_d(guide = "none") +  
  theme_minimal(base_size = 12)
p_right
```

###### Two panel graph saving with ggplot
```{r save two panel plot}
two_panel <- p_left + p_right + plot_annotation(tag_levels = "A")
two_panel
ggsave("fig_q2_two_panel.png", two_panel, width = 20, height = 5, dpi = 300)
```

###### 2×5 graph：FHP vs EOP
```{r Create scatter plot and save}
p_scatter <- df |>
  filter(!is.na(fhp_size_mm), !is.na(eop_size_mm), !is.na(age_group_name), !is.na(sex_name)) |>
  ggplot(aes(fhp_size_mm, eop_size_mm)) +
  geom_point(aes(color = sex_name), alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(sex_name ~ age_group_name) +            
  labs(x = "FHP size (mm)", y = "EOP size (mm)",
       title = "Association between FHP and EOP by age & sex") +
  scale_color_viridis_d(name = "sex") +
  theme_minimal(base_size = 12)
p_scatter
ggsave("fig_q2_p_scatter.png", p_scatter, width = 10, height = 6, dpi = 300)
```

# Problem 3
#### Q1
```{r create dataframe to solve the problem}
tb1 <- fhp_by_age |>
  select(age_group_name, total_number_people) |> 
  rename(total_number_of_people_report = total_number_people)

tb2 <- tibble(
  age_group_name = c("18-30","31-40","41-50","51-60","60+"),
  total_number_people = c(300, 200, 200, 200, 300)
) |>
  rename(total_number_of_people_author = total_number_people)

tb <- inner_join(tb1, tb2, by = "age_group_name") |> 
  knitr::kable()
tb
```
**Explain**: 
Age-group: Our data yield closely matching the paper’s target.

The corrected raw data shows 303 for 18–30, 204 for 31–40, 207 for 41–50, 200 for 51–60, and 305 for 60+. This nearly matches the authors' allocation, with only slight changes. 

Age and gender distribution is equally split: 18-30, 31-40, 41-50, 51-60, 60+, consistent with the statement balance.

#### Q1
```{r create a dataframe show difference}
tb3 <- sex_mean_sd |>
  select(sex_name, mean, sd) |> 
  rename(report_mean = mean) |> 
  rename(report_sd = sd)

tb4 <- tibble(
  sex_name = c("female","male"),
  author_mean = c(24, 28),
  author_sd = c(11, 15)
) 

tb5 <- inner_join(tb3, tb4, by = "sex_name") |> 
  rename(sex = sex_name) |>
  mutate(difference_mean = report_mean - author_mean,
         difference_sd = report_sd - author_sd) |> 
  knitr::kable()
tb5
```
**Explain**: 
The authors report gender-specific FHP measurements: 28 ± 15 mm for males and 24 ± 11 mm for females. The data yields: men 28.51 ± 14.67 mm, females 23.73 ± 10.62 mm, and overall 26.10 ± 13.01 mm. These figures agree with the authors' findings in size and direction, with rounding/sample cleaning margins.

FHP (mean±SD): Males ≈ `r nsr_df |> filter(sex_name=="male") |> summarize(m=mean(fhp_size_mm,na.rm=T), s=sd(fhp_size_mm,na.rm=T)) |> mutate(across(everything(), ~round(.x,2))) |> paste0(collapse=", ") `mm; females ≈ `r nsr_df |> filter(sex_name=="female") |> summarize(m=mean(fhp_size_mm,na.rm=T), s=sd(fhp_size_mm,na.rm=T)) |> mutate(across(everything(), ~round(.x,2))) |> paste0(collapse=", ")` mm—aligned with ~28±15 and ~24±11.

#### Q2
```{r show total prevalence}
df |> 
  filter(!is.na(eop_enlarged)) |>
  summarise(
    n = sum(!is.na(eop_size_mm)),
    cases = sum(eop_enlarged == "enlarged"),
    prev = round(cases / n, 4)
)
df
```
**Explain**: 
This study defines EEOP as EOP size groups 2-5 (≥10 mm) as “enlarged.” 

Categorical size: {2,3,4,5};

Continuous criterion: eop_size_mm > 10.

EEOP prevalence: 
The 393 “enlarged” instances with prevalence of `r nsr_df |> mutate(enlarged = eop_size_mm %in% 2:5) |> summarize(p = mean(enlarged, na.rm=TRUE)) |> pull(p) |> scales::percent(accuracy=0.1)`, matching with "33%," suggesting agreement.


#### Q3
```{r show the 60+ age group frequency}
nsr_df |> 
  filter(!is.na(age_group_name), !is.na(fhp_size_mm)) |> 
  summarise(
    n = n(),                                
    cases = sum(fhp_size_mm > 40),              
    fre  = cases / n * 100, 
    .by = age_group_name) |> 
  filter(age_group_name == "60+")
```
**Explain**: 
The prevalence of FHP > 40 mm age group: 
18–30 years: 6.6%, 
31–40 years: 5.9%, 
41–50 years: 8.7%, 
51–60 years: 11.0%, 
60+ years: `r nsr_df |> filter(age_group_name=="60+") |> summarize(p=mean(fhp_size_mm>40, na.rm=TRUE)) |> pull(p) |> scales::percent(accuracy=0.1)`
This aligns with 34.5% 

## Summary
The new data reflect the paper's results that FHP and EEOP prevalence increase with age and male sex. The study is cross-sectional with quota sampling and no direct phone-use exposure; counts should be rates; distributional information was underreported. Data show correlations, not causation. The “phones cause horn growth” claim needs individual-level exposure, longitudinal or interventional designs, robust confounding control, and measurement-reliability studies.
